The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-04-17 VegOzone.tex
Some more text for indroduction.

\introduction  %% \introduction[modified heading if necessary]
\label{sec:intro}
\begin{itemize}
\item ozone in general
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves.Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item metrics of ozone: aot40 and pod
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak {\bf TODO: Cite review article - find newer articles.}
  \end{itemize}
\item future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
\item meteorological situation 2018 vs 2019
\end{itemize}

@@

A climatology derived for global irradiance $Q_0$ is shown in Fig.~\ref{fig:global_rad_clim}a) for both, climatological average of hourly maximum $\left<Q_0^\mathrm{max}\right>$, minimum $\left<Q_0^\mathrm{min}\right>$, and mean $\left<Q_0^\mathrm{mean}\right>$ irradiance observed at Svanvik. In agreement with the geometrically derived sunshine duration shown in Fig.~\ref{fig:sunlight_fennoscandia}, $\left<Q_0^\mathrm{max}\right>$ approaches zero in December/January commonly known as polar night and reaches its maximum in June/July commonly referred to as midnight sun conditions. The highest observed $\left<Q_0^\mathrm{max}\right>$ is about $800\,\unit{W\,m^{-2}}$, while on average $\left<Q_0^\mathrm{mean}\right>$ it is about $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the climatological minimum $\left<Q_0^\mathrm{min}\right>$ lies at about $200\,\unit{W\,m^{-2}}$.

@@

The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Fig.~\ref{fig:density_distribution}a, b), hence we can put those data series together to derive a general ozone climatology for northern Fennoscandia. The correlation between Svanvik and Esrange and Pallas is rather low, 0.4 and 0.6, respectively. Thus, Svanvik cannot directly be represented by any measurements taken at either Esrange or Pallas.

@@

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $8.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring coincident with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} coincides with the estimated onset of the greening season based on the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively (day of the year 110--150 at Svanvik, Fig~\ref{fig:greening_season_change_Svanvik}).

The density distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. The spring peak apparently occurs slightly earlier (in the beginning of April). As Svanvik is located lower than the other sites, the earlier ozone peak may point to a strong interplay between ozone uptake by vegetation after the onset of greening, for at higher altitude plants sprout later. The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during night time.

@@

Finally, we shall look at the comparison of climatologies derived from observational ozone data and the various reanalysis products (Fig.~\ref{fig:ozone_climatologies}). As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis (Fig.~\ref{fig:ozone_climatologies}a) is in general too low ($(9\pm 7)\,\unit{ppb}$) and displays a erroneous seasonal cycle. MACC reanalysis ozone concentrations peak already in March followed by a second peak in July. Also CAMSRA matches the observed ozone climatology hardly (Fig.~\ref{fig:ozone_climatologies}b). Despite displaying only one ozone peak, it still does not reproduce the seasonality in northern Fennoscandia at all. The CAMSRA spring peak appears to lag behind by $1\,\unit{month}$ and is $5\,\unit{ppb}$ lower, whereas the minimum occurs in January compared to August/September. The annual amplitude ($(26\pm 1)\,\unit{ppb}$) is larger than in the climatology derived from observations ($19\,\unit{ppb}$). Modeled \chem{[O_3]} at Svanvik are even slightly higher than at Pallas. In contrast, the ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia quite well (Fig.~\ref{fig:ozone_climatologies}c), but is on average still slightly too low ($(2.8\pm 0.5)\,\unit{ppb}$).

@@

In Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals}, the probability density functions of daily mean ozone concentration residuals for Esrange and Pallas are shown. Both stations display a clear deviation from the climatological mean, however, not significant. On average, \chem{[O_3]} had been elevated by $(1.9\pm 5.7)\,\unit{ppb}$ and $(1.9\pm 5.4)\,\unit{ppb}$, respectively. Both distributions are slightly skewed towards positive deviations with $\frac{1}{4}$ of days (3. quantile, q3) falling above $5.8\,\unit{ppb}$ and $5.6\,\unit{ppb}$, respectively.

Figure~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik} displays the probability density functions of daily mean ozone concentration residuals for Svanvik based on data taken in 2018/19 with respect to the climatology derived for northern Fennoscandia (denoted NF in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}a) and the climatology derived for Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}b).
As previously mentioned, \chem{\left<[O_3]\right>} at Svanvik are $(6.6\pm 7.1)\,\unit{ppb}$ lower than the northern Fennoscandic climatology. For 2018 and 2019, we find $\Delta\chem{\left<[O_3]\right>} = (-3.8\pm 5.6)\,\unit{ppb}$ and $\Delta\chem{\left<[O_3]\right>} = (-5.2\pm 5.0)\,\unit{ppb}$, respectively. While the probability density of the residuals between the two climatologies is skewed towards negative values, the skew of 2018/19 is positive.
With respect to the climatology for Svanvik (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}b), the probability density functions average for 2018/19 lies at $\Delta\chem{\left<[O_3]\right>} = (2.6\pm 5.8)\,\unit{ppb}$ and $\Delta\chem{\left<[O_3]\right>} = (1.2\pm 5.0)\,\unit{ppb}$, respectively.
Neither of the \chem{\Delta\left<[O_3]\right>} is significant by statistical means, but daily average ozone concentrations are slightly elevated in 2018 compared to 2019.

The \chem{\left<\Delta\left<[O_3]\right>\right>} derived for Svanvik in 2018 is slightly larger than for Esrange and Pallas. This could mean either that 2018 had been more exceptional at Svanvik or the Svanvik climatology may need a correction for increased tropospheric background ozone concentration. Based on the meteorological conditions, we may assume 2019 as an average year which would lead to a deduction of a constant bias of about $1.2\,\unit{ppb}$ in the Svanvik climatology. Daily \chem{\Delta_{present-past}\left<[O_3]\right>} from Esrange and Pallas, wherein present is defined as the average of 2010--2018 and past as the respective beginning of data taking until 1997, do not reveal a clear trend: We find for Esrange $(1.5\pm 3.1)\,\unit{ppb}$ and for Pallas $(-2.5\pm 3.7)\,\unit{ppb}$. Former is compatible with our assumed bias for the Svanvik climatology.

To further investigate the ozone residuals, we conduct a Student's~t-test ($t = \frac{\chem{\Delta[O_3]}}{\sigma}$), assuming the same sample uncertainty $\sigma$ in both, climatology and the year in question. Results for Esrange and Pallas are shown in Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test}. This t-test clearly shows, that at both sites for at least $50\,\unit{\%}$ of days the residuals are significant on the $1\sigma$ level. In Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}, the test statistics for Svanvik are displayed. From Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}a), we can deduce, that the residuals of the Svanvik climatology from the northern Fennoscandic climatology are significant on the $2\sigma$ level for at least half of the days, while the residuals are less certain in 2018 and 2019. In Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}b) the same is shown for the years 2018/19 with respect to the climatology from Svanvik. We find that in less than $50\,\unit{\%}$ of days the ozone concentration residuals are significant on the $1\sigma$ level.

In accordance to the above findings, we adjust the derived ozone climatology for Svanvik by adding a constant bias correction factor $\Delta\chem{[O_3]_\mathrm{bias-corr}} = 1.2\,\unit{ppb}$ and estimate the ozone concentration anomalies for 2018/19. In Figure~\ref{fig:ozone_signific}, the resulting percentage of days above/below $\pm 1\,\sigma$ from the bias-corrected climatological mean is displayed for each month. Negative deviations from climatology are shown as negative percentage. Note that missing bars for certain months are both due to data availability (see Section~\ref{subsubsec:atmo_svanvik} for details about the time range of data taking at Svanhovd) and anomalies not meeting the criteria of $\pm 1\sigma$ deviation from climatological mean. In April/May 2018, ozone was significantly enhanced, while 2019 was generally an average year. In both years ozone concentrations were significantly below average in November.

@@

\section{DO3SE modeling}
To investigate which of the evironmental factors lead to the observed ozone damage on clover and tobacco {\bf TODO: Ane confirm!} at Svanhovd, we have modeled the ozone uptake using the DO3SE model.

>>>>

\introduction  %% \introduction[modified heading if necessary]
\label{sec:intro}

{\bf OBS: The following is just a copy of the 2018 dry deposition paper. It has to be substantially changed to meet the no-copycat-criteria. TODO: Rewrite in context of this article.}

Ozone is an important trace gas for all lifeforms on Earth. Depending on the place of its occurrence it has either a positive or negative connotation. In the stratosphere, ozone absorbs most of the ultraviolet (UV)-light from the sun within the range of 100--315\,\unit{nm}, thus shielding the Earth's surface from the most harmful UV-radiation. In addition, ozone is a potent greenhouse gas in both, stratosphere and troposphere. With a radiative forcing of $0.40 \pm 0.20\,\unit{W\,m^{-2}}$, it is placed third, only surpassed by \chem{CO_2} and \chem{CH_4} \citep[Chapter 8]{IPCC2013}.\\
In the troposphere and in particular in ambient air, ozone is considered as a highly toxic pollutant. Since the industrial revolution, tropospheric background ozone concentrations have been increasing in the northern hemisphere \citep[Chapter 2]{IPCC2013}. In recent years, the number of episodes of peak concentrations has been, in general, decreasing in North America and Europe due to the implementation of air quality regulations \citep[e.g.,][]{ESA:Fleming2018, ESA:Mills2018}. At the same time, fast developing countries, like e.g., China or India, saw a significant increase in ozone related air pollution. Continuously high concentrations of ambient air ozone are hazardous to the whole ecosystem. It is estimated that ozone is cause to an increase in pre-mature deaths \citep{WHO2008}, an average global loss of yield in the four major crops (wheat, rice, maize, and soybean) of about $3-15\,\unit{\%}$ \citep{PJ:Ainsworth2017} as well as $7\,\unit{\%}$ loss in primary production in forestry \citep{GCB:Wittig2009,EP:Matyssek2012}. The impact of ozone damage on vegetation and agricultural plants may affect food security in the future especially in Asia \citep{GCB:Tang2013,NCC:Tai2014,AE:Chuwah2015,GCB:Mills2018} and might be an important additional feedback to climate change \citep{Nat:Sitch2007}.\\
Elevated ozone levels at a site may originate from both, the local production of ozone from its precursors, which are transported, and from advection of ozone itself. Long-range ozone transport occurs regularly and might be most important in regions that otherwise lack precursors. Tropospheric ozone is produced in complex photochemical cycles involving precursor gases such as carbon monoxide (\chem{CO}) or volatile organic substances (VOCs -- also known as hydrocarbons) in the presents of nitrogen oxides (\chem{NO_x}). A typical reaction mechanism for \chem{CO} is sketched in the following. In a sequence of rapid reactions a peroxyl radical \chem{HO_2^\bullet} is formed through an initial reaction of \chem{CO} with a hydroxyl radical \chem{^\bullet OH}.
Via a reaction between \chem{HO_2^\bullet} and \chem{NO}, \chem{NO_2} is formed which is then photolyzed. The resulting atomic oxygen reacts then with \chem{O_2} (and also under the presence of available co-reactants) to form an ozone molecule.

Since ozone is highly reactive, its global mean life-time in the troposphere is roughly $22$ days but ranges between a few days in the tropical boundary layer to up to $1$ year in the upper troposphere \citep{JGR:Stevenson2005,ACP:Young2013}. The abundance of tropospheric ozone therefore varies, e.g., with time of the day, season, altitude, location \citep{ACP:Schnell2015}, or weather conditions in general \citep{ACP:Otero2018}. Typical concentrations of surface ozone range from $10\,\unit{ppb}$ over the tropical Pacific to $100\,\unit{ppb}$ in the downwind areas of highly emitting sources \citep[Chapter 8]{IPCC2013}. This variability poses a challenge on both, trend analysis from observations as well as validation and inter-comparison of models. From the observational side, the number of long-term observations (started before the 1950s) is limited and restricted to mainly European sites. Most of these indicate a doubling of tropospheric ozone since the 1950s \citep[Chapter 2]{IPCC2013}.

\begin{itemize}
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves.Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item metrics of ozone
  \begin{equation}
    \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
  \end{equation}
  with the ambient ozone concentration $\chem{[O_3]}$ and the concentration threshold $\mathrm{x}$. A typical value for $\mathrm{x}=40\,\unit{ppb}$.
  \begin{equation}
    \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{O_3}-y)}\cdot \mathrm{dt}},
  \end{equation}
   with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]}$ and the flux threshold $\mathrm{y}$.
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak {\bf TODO: Cite review article - find newer articles.}
  \end{itemize}
\item future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
\item meteorological situation 2018 vs 2019
\end{itemize}

@@

A climatology derived for global irradiance $Q_0$ is shown in Appendix Fig.~\ref{fig:global_rad_clim}a) for both, climatological average of hourly maximum $\left<Q_0^\mathrm{max}\right>$, minimum $\left<Q_0^\mathrm{min}\right>$, and mean $\left<Q_0^\mathrm{mean}\right>$ irradiance observed at Svanvik. In agreement with the geometrically derived sunshine duration shown in Appendix Fig.~\ref{fig:sunlight_fennoscandia}, $\left<Q_0^\mathrm{max}\right>$ approaches zero in December/January commonly known as polar night and reaches its maximum in June/July commonly referred to as midnight sun conditions. The highest observed $\left<Q_0^\mathrm{max}\right>$ is about $800\,\unit{W\,m^{-2}}$, while on average $\left<Q_0^\mathrm{mean}\right>$ it is about $450\,\unit{W\,m^{-2}}$ (regularly reached in May--July). The maximum of the climatological minimum $\left<Q_0^\mathrm{min}\right>$ lies at about $200\,\unit{W\,m^{-2}}$.

@@

The correlation of \chem{[O_3]} between Esrange, Pallas, and Jergul/Karasjok is fairly high ($r^2\approx 0.8$, Appendix Fig.~\ref{fig:density_distribution}a, b), hence we can put those data series together to derive a general ozone climatology for northern Fennoscandia. The correlation between Svanvik and Esrange and Pallas is rather low, 0.4 and 0.6, respectively. Thus, Svanvik cannot directly be represented by any measurements taken at either Esrange or Pallas.

@@

From the northern Fennoscandic climatology, one can deduce several features. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} peaks about $20\,\unit{days}$ earlier than the \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}}. With respect to the periodicity of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, it takes about $8.5\,\unit{months}$ from minimum concentration to maximum concentration, while from maximum to minimum it only takes about $4.5\,\unit{months}$. The maximum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} is at $52\,\unit{ppb}$ in late April and the minimum of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, min}} occurs at $15\,\unit{ppb}$ in the beginning of September. The steep rise of \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} in spring coincident with increasing temperatures as well as increase in global irradiance. The decrease in \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}} coincides with the estimated onset of the greening season based on the agricultural rule of thumb that the daily average temperature has to be above $5\,\unit{^\circ C}$ for $5\,\unit{days}$ consecutively (day of the year 110--150 at Svanvik, Appendix Fig~\ref{fig:greening_season_change_Svanvik}).

The density distribution for Svanvik and the respective \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, mean}}, and \chem{\left<[O_3]\right>_\mathrm{10\,\unit{d}, max}} (Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_spline}b) display a similar pattern as the northern Fennoscandic climatology. The spring peak apparently occurs slightly earlier (in the beginning of April). As Svanvik is located at lower altitude than the other sites, the earlier ozone peak may point to a strong interplay between ozone uptake by vegetation after the onset of greening, for at higher altitude plants sprout later. {\bf TODO: Is there a way to confirm this? Actual data for the three sites? Please correct if wrong! Ane? Aud?} The average ozone concentration at Svanvik is lower than at the other sites, as already mentioned, which is in league with a decrease in tropospheric ozone with decreasing altitude. In July--September, ozone is occasionally almost completely depleted during night time.

@@

Finally, we shall look at the comparison of climatologies derived from observational ozone data and the various reanalysis products (Fig.~\ref{fig:ozone_climatologies}). As representative for the sites, the nearest neighboring grid point has been chosen. The ECMWF products do not reproduce observed surface ozone well. The MACC reanalysis (Fig.~\ref{fig:ozone_climatologies}a) is in general too low ($(9\pm 7)\,\unit{ppb}$) and displays a erroneous seasonal cycle. MACC reanalysis ozone concentrations peak already in March followed by a second peak in July. Also CAMSRA matches the observed ozone climatology poorly (Fig.~\ref{fig:ozone_climatologies}b). Despite displaying only one ozone peak, it still does not reproduce the seasonality in northern Fennoscandia at all. The CAMSRA spring peak appears to lag behind by $1\,\unit{month}$ and is $5\,\unit{ppb}$ lower, whereas the minimum occurs in January compared to August/September. The annual amplitude ($(26\pm 1)\,\unit{ppb}$) is larger than in the climatology derived from observations ($19\,\unit{ppb}$). Modeled \chem{[O_3]} at Svanvik are even slightly higher than at Pallas. In contrast, the ensemble mean of the Copernicus regional model reanalysis does reproduce the seasonal cycle in northern Fennoscandia quite well (Fig.~\ref{fig:ozone_climatologies}c), but is on average still slightly too low ($(2.8\pm 0.5)\,\unit{ppb}$).

@@

In Appendix Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals}, the probability density functions of daily mean ozone concentration residuals for Esrange and Pallas are shown. Both stations display a clear deviation from the climatological mean, however, not significant. On average, \chem{[O_3]} had been elevated by $(1.9\pm 5.7)\,\unit{ppb}$ and $(1.9\pm 5.4)\,\unit{ppb}$, respectively. Both distributions are slightly skewed towards positive deviations with $\frac{1}{4}$ of days (3. quantile, q3) falling above $5.8\,\unit{ppb}$ and $5.6\,\unit{ppb}$, respectively.

Appendix Figure~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik} displays the probability density functions of daily mean ozone concentration residuals for Svanvik based on data taken in 2018/19 with respect to the climatology derived for northern Fennoscandia (denoted NF in Appendix Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}a) and the climatology derived for Svanvik (Appendix Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}b).
As previously mentioned, \chem{\left<[O_3]\right>} at Svanvik are $(6.6\pm 7.1)\,\unit{ppb}$ lower than the northern Fennoscandic climatology. For 2018 and 2019, we find $\Delta\chem{\left<[O_3]\right>} = (-3.8\pm 5.6)\,\unit{ppb}$ and $\Delta\chem{\left<[O_3]\right>} = (-5.2\pm 5.0)\,\unit{ppb}$, respectively. While the probability density of the residuals between the two climatologies is skewed towards negative values, the skew of 2018/19 is positive.
With respect to the climatology for Svanvik (Appendix Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals-Svanvik}b), the probability density functions average for 2018/19 lies at $\Delta\chem{\left<[O_3]\right>} = (2.6\pm 5.8)\,\unit{ppb}$ and $\Delta\chem{\left<[O_3]\right>} = (1.2\pm 5.0)\,\unit{ppb}$, respectively.
Neither of the \chem{\Delta\left<[O_3]\right>} is significant by statistical means, but daily average ozone concentrations are slightly elevated in 2018 compared to 2019.

The \chem{\left<\Delta\left<[O_3]\right>\right>} for Svanvik in 2018 is slightly larger than for Esrange and Pallas. This could mean either that 2018 had been more exceptional at Svanvik or the derived climatology may need a correction for increased tropospheric background ozone concentration. Based on the meteorological conditions, we assume 2019 as average year which leads to a constant bias of about $1.2\,\unit{ppb}$. Daily \chem{\Delta_{present-past}\left<[O_3]\right>} from Esrange and Pallas, wherein present is defined as the average of 2010--2018 and past as the respective beginning of data taking until 1997, do not reveal a clear trend: We find for Esrange $(1.5\pm 3.1)\,\unit{ppb}$ and for Pallas $(-2.5\pm 3.7)\,\unit{ppb}$. Former is compatible with our assumed bias for the climatology derived for Svanvik.

To further investigate the ozone residuals, we conduct a Student's~t-test ($t = \frac{\chem{\Delta[O_3]}}{\sigma}$), assuming the same sample uncertainty $\sigma$ in both, climatology and the year in question. Results for Esrange and Pallas are shown in Appendix Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test}. This t-test clearly shows, that at both sites the residuals are significant on the $1\sigma$ level for at least $50\,\unit{\%}$. In Appendix Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}, the test statistics for Svanvik are displayed. From Appendix Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}a), we can deduce, that the residuals between the Svanvik climatology and the northern Fennoscandic climatology are significant on the $2\sigma$ level for at least half of the data, while the residuals are less certain in 2018 and 2019. In Appendix Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_test-Svanvik}b) the same is shown for the years 2018/19 with respect to the climatology from Svanvik. We find that less than $50\,\unit{\%}$ of ozone concentration residuals are significant on the $1\sigma$ level.

In accordance to the above findings, we adjust the derived ozone climatology for Svanvik by adding a constant bias correction factor $\Delta\chem{[O_3]_\mathrm{bias-corr}} = 1.2\,\unit{ppb}$ and estimate the ozone concentration anomalies for 2018/19. In Fig.~\ref{fig:ozone_signific}, the resulting percentage of days above/below $\pm 1\,\sigma$ from the bias-corrected climatological mean is displayed for each month. Negative deviations from climatology are shown as negative percentage. Note that missing bars for certain months are both due to data availability (see Section~\ref{subsubsec:atmo_svanvik} for details about the time range of data taking at Svanhovd) and anomalies not meeting the criteria of $\pm 1\sigma$ deviation from climatological mean. In April/May 2018, ozone was significantly enhanced, while 2019 was generally an average year. In both years, ozone concentrations were significantly below average in November.

@@

\section{DO3SE modeling}
To investigate which of the environmental factors lead to the observed ozone damage on clover and tobacco {\bf TODO: Ane confirm!} at Svanhovd, we have modeled the ozone uptake using the DO3SE model.
